---
title: "Prescribing in Grampian"
output: html_document
---

```{r, message = F, echo = F, warning = F}
knitr::opts_chunk$set(message = F, echo = F, warning = F)
```


```{r}
library(dplyr)
library(gt)

#load data cleaned in data_prep.R
#remove NAs - these make up a substantial cost 
grampian <- readr::read_csv("grampian_rx.csv") %>%
  filter(!is.na(bnf_product))

#BNF data 
bnf <-
  readr::read_csv("20220905_1662367590394_BNF_Code_Information.csv") %>%
  janitor::clean_names() %>%
  mutate(bnf_product = 
           ifelse(bnf_product == "DUMMY PRODUCT 21480000101",
                  "FreeStyle Libre 2", bnf_product),
         bnf_product = 
           ifelse(bnf_product == "DUMMY PRODUCT 21480000100",
                  "FreeStyle Libre", bnf_product))
```

<br> 

### Which drugs are prescribed most?
```{r}
rx_summary_table <-
  grampian %>%
  group_by(BNFItemCode) %>%
  summarise(
    total_rx = sum(NumberOfPaidItems)) %>%
  arrange(desc(total_rx)) %>%
   filter(total_rx >= 75000) %>%
  left_join(., bnf, 
            by = c("BNFItemCode" = "bnf_presentation_code")) %>%
  select(bnf_product, total_rx, bnf_section)

rx_summary_table %>%
  gt() %>%
  cols_label(bnf_product = "Drug",
             total_rx = "Annual Rx",
             bnf_section = "Drug Class") %>%
  fmt_number(total_rx, 
             decimals = 0,
             use_seps = T) %>%
  tab_options(table.align = "left",
              data_row.padding = 0,
              column_labels.font.weight = "bold")
```

<br> 

### Which drugs cost the most?
```{r}
cost_summary_table <-
  grampian %>%
  group_by(BNFItemCode) %>%
  summarise(
    total_cost = sum(GrossIngredientCost),
    total_rx = sum(NumberOfPaidItems),
    cost_rx =
      total_cost /
      total_rx) %>%
  arrange(desc(total_cost)) %>%
   filter(total_cost >= 500000) %>%
  left_join(., bnf, 
            by = c("BNFItemCode" = "bnf_presentation_code")) %>%
  select(bnf_product, total_cost, total_rx, cost_rx, bnf_section)

cost_summary_table %>%
  gt() %>%
  cols_label(bnf_product = "Drug",
             total_cost = "Annual Cost",
             total_rx = "Annual Rx",
             cost_rx = "Cost per Rx",
             bnf_section = "Drug Class") %>%
  fmt_currency(c(total_cost, cost_rx),
               decimals = 0,
               currency = "GBP") %>%
  fmt_number(total_rx, 
             decimals = 0,
             use_seps = T) %>%
  tab_options(table.align = "left",
              data_row.padding = 0,
              column_labels.font.weight = "bold")
```


<br>  

### Which classes of drugs cost the most?
```{r}
class_cost_table <-
  grampian %>%
  group_by(bnf_section) %>%
  summarise(
    total_cost = sum(GrossIngredientCost),
    total_rx = sum(NumberOfPaidItems),
    cost_rx =
      total_cost /
      total_rx) %>%
  arrange(desc(total_cost)) %>%
   filter(total_cost >= 1000000) 

class_cost_table %>%
  gt() %>%
  cols_label(bnf_section = "Drug Class",
             total_cost = "Annual Cost",
             total_rx = "Annual Rx",
             cost_rx = "Cost per Rx") %>%
  fmt_currency(c(total_cost, cost_rx),
               decimals = 0,
               currency = "GBP") %>%
  fmt_number(total_rx, 
             decimals = 0,
             use_seps = T) %>%
  tab_options(table.align = "left",
              data_row.padding = 0,
              column_labels.font.weight = "bold")
```

<br> 

### Data and Methods  
This analysis includes all prescriptions in NHS Grampian dispensed outside of hospitals from July 2021 through June 2022.  

All data are available here https://github.com/JessButler/prescriptions https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community

For each prescription, details from the BNF were added from here https://applications.nhsbsa.nhs.uk/infosystems/data/showDataSelector.do?reportId=126

Number of prescriptions and costs per item calculated as specified here https://www.isdscotland.org/Health-topics/Prescribing-and-medicines/_docs/Open_Data_Glossary_of_Terms.pdf?1 

All code to produce this report is here https://github.com/JessButler/prescriptions

Any questions or comments please email JessicaButler@abdn.ac.uk
